Exploring environmental impacts of colonial contacts on Indigenous communities in northeastern Taiwan

### 2025-02-12 analysis/paper/paper.qmd
add paragraphs

Given that deerskin was a primary commodity in local regional exchange since the 14th century and was in high demand during the Dutch colonization period [@Cheng2017], the skeletal compositions, fragmentation, and butchery patterns of deer bones were examined. The distribution of body portions and butcher patterns can shed light on how indigenous people utilize deer in response to economic demands during the foreign presence. The proportion of deer skeletal elements was quantified using NISP and classified into groups based on the meat yield: meaty elements (e.g., upper limb), mid-meaty elements (e.g., body parts), and non-meaty elements (e.g., skull bones and lower limb) [@Madrigal2004]. While lower limbs and skull bones have minimal meat, they are often valued in many cultures for marrow and soft tissues, such as tongue and brain. Thus, non-meaty elements were considered as secondary source of nutrition in the interpretation. Similarly, the degree of fragmentation and the weight of broken unidentified bones could also imply the extraction of marrow or grease. The degree of fragmentation is calculated by dividing Minimal Number of Elements (MNE) by NISP. For butcher patterns, cut marks on major limb joints, including the shoulder, elbow, wrist, hip, knee, and ankle [@Lyman2005], were identified and quantified for a comparison between different temporal sequences. In addition, modifications such as cut marks, fracturing, and burning were examined, especially the cut marks associated with the skinning process. Skinning processes can be identified by the encircling and straight marks left on the lower limb, such as metapodials, which result from the series of cuts typically made to acquire intact animal hides [@Costamagno2019; @Isaakidou2007].

@@

For animal classes, the relative abundance shows that mammals are the primary dietary source throughout all layers, with fish as the secondary source in layer 4 and birds in layer 3 and layer 2 (@fig-NISP-proportion: A). Fish and birds are both identified from layer 5 to layer 3; however, their proportions (less than 10%) are much lower compared to mammals. The reptile is represented by one piece of turtle shell found in layer 4, which might not be a common food content at the site. Within those identifiable mammals, deer is the major source from layer 6 to layer 2 that accounts for at least 85% per layer, while layer 1 is predominated only by cattle (@fig-NISP-proportion: B). Wild boar was present from layer 6 to layer 3, and had a higher proportion in layer 4 (`r CL4_boar` %), while muntjac appeared in layer 5, and was continuously present until layer 2. In general, more diverse animals were found from layer 5 to layer 3, especially in layer 4, where boar and fish have slightly higher proportions compared with other layers. Given that the taxonomic richness, or diversity of animals, could be impacted by sample size varied from layer to layer, the relationship between the number of taxa and sample size (NISP) was examined by linear regression model. The result shows no correlation between the diversity of animals and sample size ($R^2$ = `r div_lm[1]`, P = `r div_lm[2]`), which rules out the potential sample size bias.

@@

```{r}
#| label: tbl-ubiquity
#| echo: false
#| tbl-cap: "Ubiquity of faunal categories over the 40 units"
#| tbl-cap-location: bottom
# calculate the number of units for each cultural layers
kwl_unit_per_period <-
  kwl_chro_6 %>%
  mutate(all = pmap_chr(select(., -Pit, -Grid), ~toString(unique(na.omit(c(...)))))) %>%  # unite and remove duplicates
  mutate(unit = str_extract(Pit, "[A-Z]+[0-9]{3}")) %>%
  distinct(unit, .keep_all = TRUE) %>%
  mutate(CL1 = ifelse(str_detect(all, "CL1"), unit, NA)) %>%
  mutate(CL2 = ifelse(str_detect(all, "CL2"), unit, NA)) %>%
  mutate(CL3 = ifelse(str_detect(all, "CL3"), unit, NA)) %>%
  mutate(CL4 = ifelse(str_detect(all, "CL4"), unit, NA)) %>%
  mutate(CL5 = ifelse(str_detect(all, "CL5"), unit, NA)) %>%
  mutate(CL6 = ifelse(str_detect(all, "CL6"), unit, NA))

@@

@@

@tbl-ubiquity shows how we can have a caption and cross-reference for a plot. Note that figure label and cross-references must both be prefixed with `fig-`

```{r}
#| label: demo-inline-code
#| echo: false
x <- round(pi, 2)
```

Here is an example of inline code `r x` in the middle of a sentence.

@@

>>>>

Given that deerskin was a primary commodity in local regional exchange since the 14th century and was in high demand during the Dutch colonization period [@Cheng2017], the skeletal compositions, fragmentation, and butchery patterns of deer bones were examined. The distribution of body portions and butcher patterns can shed light on how indigenous people utilize deer in response to economic demands during the foreign presence. The proportion of deer skeletal elements was quantified using NISP and classified into groups based on the meat yield: meaty elements (e.g., upper forelimb and upper hindlimb ), mid-meaty elements (e.g., body parts), and less-meaty elements (e.g., skull bones and lower limb) [@Madrigal2004]. While lower limbs and skull bones have minimal meat, they are often valued in many cultures for marrow and soft tissues, such as tongue and brain. Thus, non-meaty elements were considered as secondary source of nutrition in the interpretation. Similarly, the degree of fragmentation could also imply the extraction of marrow or grease. The degree of fragmentation is calculated by dividing Minimal Number of Elements (MNE) by NISP. For butcher patterns, cut marks on major limb joints, including the shoulder, elbow, wrist, hip, knee, and ankle [@Lyman2005], were identified and quantified for a comparison between different temporal sequences. In addition, modifications such as cut marks, fracturing, and burning were examined, especially the cut marks associated with the skinning process. Skinning processes can be identified by the encircling and straight marks left on the lower limb, such as metapodials, which result from the series of cuts typically made to acquire intact animal hides [@Costamagno2019; @Isaakidou2007].

@@

For animal classes, the relative abundance shows that mammals are the primary dietary source throughout all layers, with fish as the secondary source in CL4 and birds in CL3 and CL2 (@fig-NISP-proportion: A). Fish and birds are both identified from CL5 to CL3; however, their proportions (less than 10%) are much lower compared to mammals. The reptile is represented by one piece of turtle shell found in CL4, which might not be a common food content at the site. Within those identifiable mammals, deer is the major source from CL6 to CL2 that accounts for at least 85% per layer, while CL1 is predominated only by cattle (@fig-NISP-proportion: B). Wild boar was present from CL6 to CL3, and had a higher proportion in CL4 (`r CL4_boar` %), while muntjac appeared in CL5, and was continuously present until CL2. In general, there is an increase in animal diversity in CL4 and CL5, especially in CL4, where boar and fish have slightly higher proportions compared with other layers. Given that the taxonomic richness, or animal diversity, could be impacted by sample size varied from layer to layer, the relationship between the number of taxa and sample size (NISP) was examined by linear regression model. The result shows no correlation between the diversity of animals and sample size ($R^2$ = `r div_lm[1]`, P = `r div_lm[2]`), which rules out the potential sample size bias.

@@

```{r}
#| label: tbl-ubiquity
#| echo: false
#| tbl-cap: "Ubiquity of faunal categories across the sampling area at Kiwulan."
#| tbl-cap-location: bottom
# calculate the number of units for each cultural layers
kwl_unit_per_period <-
  kwl_chro_6 %>%
  mutate(all = pmap_chr(select(., -Pit, -Grid), ~toString(unique(na.omit(c(...)))))) %>%  # unite and remove duplicates
  mutate(unit = str_extract(Pit, "[A-Z]+[0-9]{3}")) %>%
  distinct(unit, .keep_all = TRUE) %>%
  mutate(CL1 = ifelse(str_detect(all, "CL1"), unit, NA)) %>%
  mutate(CL2 = ifelse(str_detect(all, "CL2"), unit, NA)) %>%
  mutate(CL3 = ifelse(str_detect(all, "CL3"), unit, NA)) %>%
  mutate(CL4 = ifelse(str_detect(all, "CL4"), unit, NA)) %>%
  mutate(CL5 = ifelse(str_detect(all, "CL5"), unit, NA)) %>%
  mutate(CL6 = ifelse(str_detect(all, "CL6"), unit, NA))

@@

CL4_ubi_deer <- fauna_ubiquity$CL4[1]
CL4_ubi_boar <- fauna_ubiquity$CL4[2]
CL4_ubi_munt <- fauna_ubiquity$CL4[3]

@@

@tbl-ubiquity shows deer had been the most frequent animal from CL6 to CL2, with the highest ubiquity in CL4 and CL5 (`r CL4_ubi_deer`), but disappeared in CL1. The second and third frequent animal are boar and muntjac. Although both animals reached their highest ubiquity in CL4 (`r CL4_ubi_boar` for boar, and `r CL4_ubi_munt` for muntjac), their presence patterns are slightly different. Boar was utilized as early as deer in CL6, declined in CL3, and disappeared in CL2, whereas muntjac showed a cultural lag, appearing one phase later than boar. Cattle had a low ubiquity in CL5 and high ubiquity in CL6. Small animals (bird, fish, and turtle) had much lower ubiquity than large animals and they had a slightly increase in ubiquity in CL3. The result of animal ubiquity correspond with the result of their quantity (NISP), suggesting that deer is the most common resource at the site, with an increase utilization of boar and muntjac in CL4.

```{r}
#| label: fig-deer-portion
#| echo: false
#| fig-cap: "The frequency of skeletal portions using NISP of deer across cultural layers at Kiwulan"

knitr::include_graphics(here::here("analysis/figures/deer_portion.png"))
```

Since deer remains was the major component of faunal assemblage at Kiwulan according to their high quantity and ubiquity, their body parts were examined to explore any changes in dietary practices and preferences.

@@

played a consistent role in subsistence strategies.
A taxon can have high NISP but low ubiquity if concentrated in a few contexts (e.g., feasting deposits).
A taxon with high ubiquity but low NISP may have been used frequently but in small amounts.

Other than quantity, the measure of presence and absence of a given taxon across the sampeld units could give insights into the frequency and distribution consistency of taxa at a settlement scale.

High-ubiquity taxa (e.g., deer, pigs) likely played a consistent role in subsistence strategies. Low-ubiquity taxa (e.g., rare birds, marine species) may indicate opportunistic hunting, ritual use, or seasonal availability. Changes in ubiquity across layers may reflect environmental shifts or human adaptations. For example, a decline in large mammal ubiquity might suggest deforestation or overhunting, leading to a shift toward smaller prey.

Issue of Taphonomic & Preservation Bias: Some taxa may be less ubiquitous due to poor bone preservation rather than actual scarcity in the past. Small and fragile bones (e.g., fish, rodents) might have low ubiquity due to taphonomic processes.
